---
title: "Mapping point data using R"
author: "Thibaut Jombart"
date: '`r format(Sys.time(), "%d %B, %Y")`'
categories: ["R"]
tags: ["R Markdown", "map", "plot"]
output: html_document
---



```{r, echo = FALSE}
options(width = 100)

knitr::opts_chunk$set(
  fig.width = 7, 
  fig.height = 7,
  dev = 'png',
  fig.path = "figs/",
  collapse = TRUE
)
```


In this tutorial, we illustrate how to generate maps using point
data. Typically, individual points will correspond to the location of a reported
case. We will illustrate several approaches including: basic point maps,
representing point density, and using spatial clustering to approximate spatial
structures.





### The data: a simulated Ebola outbreak in Freetown

This example uses the latest devel version of the
[http://www.repidemicsconsortium.org/outbreaks/](*outbreaks*) package, which can
be installed by typing:

```{r install, eval = FALSE }

devtools::install_github("reconhub/outbreaks")

```

Note that this requires `devtools` to be installed with its dependencies.


Once installed, we load the package and have a quick look at the data:


```{r data}

library(outbreaks)

class(ebola_sim)
names(ebola_sim)

head(ebola_sim$linelist)

## column 10 = longitude
## column 11 = latitude
lonlat <- ebola_sim$linelist[, c(10, 11)]
head(lonlat)
nrow(lonlat)

```

`ebola_sim` is a `list` whose first component is a `data.frame` containing the
**linelist** of a simulated outbreak. Each row is a different reported case,
with individual spatial coordinates provided as longitude and latitude. We
isolate these coordinates in a 2-columns `data.frame`.

**Tip to use your own data**: you can read data from a spreadsheet into R and
use the procedure described here, making sure you indicate the right columns for
longitude and latitude (in our example, columns 10 and 11). The most common
formats will be:

- `.xlsx`: use the function `read.xlsx` from the `xlsx` package
- `.csv`: use the function `read.csv` from base R
- `.txt`: use the function `read.table` from base R



### Basic point map

The packages *ggmap* and *ggplot2* can be used to visualise the distribution of
cases on a map. The function `get_map` can be tailored (see `?get_map`) to use
different zoom levels and different map backgrounds. Once the background map has
been created, normal *ggplot2* features can be used to overlay graphics on the
map:

```{r basic_map}

library(ggmap)
library(ggplot2)

## spatial locations
base_points <- ggplot(data = lonlat, aes(x = lon, y = lat))

## make basic map
base <- ggmap(get_map(lonlat, zoom = 13), base_layer = base_points)

## visualise map and spatial distribution
base + geom_point(col = "red", alpha = .3)

```

Here, transparency (the `alpha` parameter) is used to give an idea of point
density when close locations result in partly overlapping symbols. In the following part we see how this can be improved using *density estimation* or *spatial clustering*.




### Visualising point density

Kernel density estimation can be used to visualise the density of points (here,
a proxy for the number of cases) on a map; this can be done easily using
*ggplot2*'s `geom_density_2d` feature:

```{r density_map}

base + geom_point(col = "red", alpha = .3) + geom_density_2d()

```

The amount of smoothing used in the density estimation is determined by a
parameter `h`; for instance, for less smoothing (adding some customisation for
colors and line types):

```{r density_map_2}

base + geom_point(col = "black", alpha = .15) +
       geom_density_2d(color = "red", h = 0.005)


```



One can also represent case density using a heatmap:

```{r density_map_3}

heat_colors <- scale_fill_gradientn("Case density", colors = c("white", "gold", "red", "darkred"))

base + stat_density_2d(aes(fill = ..level..), geom = "polygon", alpha = .3) +
       heat_colors +
       theme(legend.position = c(.1,.8))
       
```




### Using spatial clustering

Point data may be non-randomly distributed, in which case it would make sense to
identify clusters of infections. This can be achieve using various
algorithms. Here, we illustrated the use of the *Kmeans* approach. This method
uses spatial coordinates and a pre-defined number of clusters to identify
spatial groups, so as to minimize the overall distance between cases and their
group centroid.

We illustrate this procedure for an arbitrary number of groups set to 60 (this
number can be varied to suit the needs of a specific dataset):

```{r clust_map}

clust <- kmeans(lonlat, centers = 60)

xy_clust <- data.frame(clust$centers, number = clust$size)

base + geom_point(data = xy_clust,
       		  aes(x = lon, y = lat, size = number, color = cluster),
       		  alpha = .5, color = "red") +
       scale_size("Number of cases", range = c(1,15)) +
       theme(legend.position = c(.1,.8))
       
```

